---
title: "Customer Churn Analysis: Telco customer churn on the IBM dataset"
author: "Aliyu Atiku Mustapha"
date: "2023-08-13"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 2
    code_folding: hide
---
# Analysis Objectives

The purpose of this analysis is to answer 5 business questions using the Telco
Customer Churn - IBM Dataset.

1. What are the main factors contributing to customer churn?

2. Can we identify patterns or behaviors that precede customer churn?

3. Which customer segments have the highest churn rates?

4. Are there specific products or services associated with higher churn rates?

5. Can we predict which customers are likely to churn in the near future?

# The Data

## Telco Customer Churn: IBM Dataset

The data is from a fictional telco company that provided home phone and Internet services to 7043 customers in California in Q3
The data has 7043 measurements of 33 different variables.

The dataset is available on
[Kaggle](https://www.kaggle.com/datasets/yeanzc/telco-customer-churn-ibm-dataset). The data has the following columns and their descriptions:

| Column Name | Column Description 
|--------- |--------- 
| CustomerID |  A unique ID that identifies each customer
| Count | A value used in reporting/dashboarding to sum up the number of customers in a filtered set
| Country | The country of the customer’s primary residence
| State | The state of the customer’s primary residence
| City | The city of the customer’s primary residence
| Zip Code | The zip code of the customer’s primary residence
|Lat Long| The combined latitude and longitude of the customer’s primary residence
|Latitude| The latitude of the customer’s primary residence
|Longitude| The longitude of the customer’s primary residence
|Gender| The customer’s gender| Male, Female
|Senior Citizen| Indicates if the customer is 65 or older - Yes or No
|Partner| Indicate if the customer has a partner - Yes or No
|Dependents| Indicates if the customer lives with any dependents - Yes or No
|Tenure Months| Indicates the total amount of months that the customer has been with the company by the end of the quarter specified above
|Phone Service| Indicates if the customer subscribes to home phone service with the company - Yes or No
|Multiple Lines| Indicates if the customer subscribes to multiple telephone lines with the company - Yes or No
|Internet Service| Indicates if the customer subscribes to Internet service with the company - No, DSL, Fiber Optic, Cable
|Online Security| Indicates if the customer subscribes to an additional online security service provided by the company - Yes or No
|Online Backup| Indicates if the customer subscribes to an additional online backup service provided by the company - Yes or No
|Device Protection| Indicates if the customer subscribes to an additional device protection plan for their Internet equipment provided by the company - Yes or No
|Tech Support| Indicates if the customer subscribes to an additional technical support plan from the company with reduced wait times - Yes or No
|Streaming TV| Indicates if the customer uses their Internet service to stream television programing from a third party provider - Yes or No
|Streaming Movies| Indicates if the customer uses their Internet service to stream movies from a third party provider - Yes or No
|Contract| Indicates the customer’s current contract type - Month-to-Month, One Year, Two Year
|Paperless Billing| Indicates if the customer has chosen paperless billing - Yes or No
|Payment Method| Indicates how the customer pays their bill - Bank Withdrawal, Credit Card, Mailed Check
|Monthly Charge| Indicates the customer’s current total monthly charge for all their services from the company
|Total Charges| Indicates the customer’s total charges, calculated to the end of the quarter specified above
|Churn Label| Yes = the customer left the company this quarter No = the customer remained with the company Directly related to Churn Value
|Churn Value| 1 = the customer left the company this quarter 0 = the customer remained with the company Directly related to Churn Label
|Churn Score| A value from 0-100 that is calculated using the predictive tool IBM SPSS Modeler The model incorporates multiple factors known to cause churn The higher the score, the more likely the customer will churn
|CLTV| Customer Lifetime Value A predicted CLTV is calculated using corporate formulas and existing data The higher the value, the more valuable the customer High value customers should be monitored for churn
|Churn Reason| A customer’s specific reason for leaving the company Directly related to Churn Category

```{r warning=FALSE, setup, message=FALSE}
# Load tidyverse package for data analysis
library(tidyverse)
# Load readxl package for reading excel file
library(readxl)
# Load rsample package for modeling
library(rsample)
# Load naniar package for imputation
library(naniar)
# Load corrplot package for correlation plot
library(corrplot)
# Load gridExtra package for combining plots
library(gridExtra)
# Load gbm package for Logistic regression modeling
library(gbm)
# Load randomForest package for Logistic regression modeling
library(randomForest)
# Load rpart package for Decision Trees modeling
library(rpart)
# Load pROC package for model performance measuring
library(pROC)
# Load ROCR package for model performance measuring
library(ROCR)
# Load DT package for table display
library(DT)
# Load knitr package 
library(knitr)
```

## Load Telco Data

```{r, read_data_file, message=FALSE, warning=FALSE}
# load the data set
telco_data <- read_excel("E:/My Data Science Portfolio/Datasets/Telco_customer_churn.xlsx")
# Calculate total number of rows in data
cat("The Telco dataset has",nrow(telco_data), "rows.")
# Display the first 500 rows of imported data
datatable(
  telco_data[1:500,],
  options = list(pageLength = 20),
  caption = "The first 500 rows of loaded data.",
  fillContainer = T)
```

## Data Properties

To understand the data composition, the class of each variable together with
the number and proportion of missing values for each variable will provide a 
deeper insight to the data structure and how it could be useful for analysis.

```{r data_str, message=FALSE, warning=FALSE}
# Check for class of each variable
class_table <- sapply(telco_data, class)
class_table <- data.frame(
  Variable = names(class_table),
  Class = as.character(class_table),
  stringsAsFactors = FALSE
)
# Check for the proportion of missing values in full data
x <- data.frame(miss = colSums(is.na(telco_data)),
                prop = round((colSums(is.na(telco_data)) 
                              / nrow(telco_data)) * 100,
                          2))
properties <- cbind(class_table,
                    Values_missing = x[, 1],
                    Proportion_missing = x[, 2])
# Display the properties of the data
datatable(
  properties,
  options = list(pageLength = 35),
  caption = "Table displaying the properties of data.",
  fillContainer = T)
```

Note: Data has 2 variables with missing values. 

The variable 'Churn Reason' has a high proportion of missing values, 73.46% and
will be dropped from the data, while the variable 'Total Charges' will be
imputed with the mean value because it has less than 0.15% missing to avoid
dropping any rows. Imputing by the mean on a small number of missing values (11
rows to be precise) will not skew the data.

## Data Processing

Some variables are not needed while some need reformatting by changing the class of variables to the appropriate type, for each variable.

```{r clean_data, message=FALSE, warning=FALSE}
# Step 1: Drop non-required variables in data
telco_cleaned <- telco_data %>%
  select(
    -CustomerID, # Does not provide any insights to churning
    -Count, # Same value no variation
    -Country, # Same value no variation
    -State, # Same value no variation
    - City, # Duplicate as Zip Codes serve same purpose
    -`Lat Long`, # Does not provide any insights to churning
    -Latitude, # Duplicate found in Lat Long
    -Longitude, # Duplicate found in Lat Long
    -`Churn Value`, # Replicate of Churn Label
    -`Churn Reason` # High proportion of missing values
  )
# Step 2: Rename variables using underscore instead of space
names(telco_cleaned) <- c("Zip_Code","Gender", "Senior_Citizen", "Partner",
                          "Dependents", "Tenure_Months", "Phone_Service",
                          "Multiple_Lines", "Internet_Service", 
                          "Online_Security", "Online_Backup", 
                          "Device_Protection", "Tech_Support", "Streaming_TV", 
                          "Streaming_Movies", "Contract", "Paperless_Billing", 
                          "Payment_Method", "Monthly_Charges", "Total_Charges", 
                          "Churn_Label", "Churn_Score", "CLTV")
# Step 3: Impute missing values for Total_Charges variable with the mean value 
telco_cleaned$Total_Charges <- impute_mean(telco_cleaned$Total_Charges)
# Step 4: Convert all categorical variables from character to factor
# Group all character variables
cat_variables <- c("Gender", "Senior_Citizen", "Partner", "Dependents",
                  "Phone_Service", "Multiple_Lines", "Internet_Service",
                  "Online_Security", "Online_Backup", "Device_Protection",
                  "Tech_Support", "Streaming_TV", "Streaming_Movies", 
                  "Contract", "Paperless_Billing", "Payment_Method",
                  "Churn_Label")
# Convert from character variables to factor variables
telco_cleaned[cat_variables] <-
  lapply(telco_cleaned[cat_variables], as.factor)

# Step 5: Encode binary variables from Yes/No to 1/0
# Group all binary variables
bin_variable <- c("Senior_Citizen", "Partner", "Dependents","Phone_Service", 
                  "Multiple_Lines", "Online_Security", "Online_Backup",
                  "Device_Protection", "Tech_Support","Streaming_TV", 
                  "Streaming_Movies", "Paperless_Billing", "Churn_Label")
# Convert from binary variables to numeric values
telco_cleaned[bin_variable] <-
  lapply(telco_cleaned[bin_variable], function(z) ifelse(z == "Yes", 1, 0))
# Display the first 500 rows of cleaned data
datatable(
  telco_cleaned[1:500,],
  options = list(pageLength = 20),
  caption = "The first 500 rows of cleaned data with reduced variables.",
  fillContainer = T)
```

# Analysis

## 1. What are the main factors contributing to customer churn?

Find out the proportion of overall customers that churn, and high value
customers that churn.

```{r, explore, message=FALSE, warning=FALSE}
# Churn Distribution
churn_rate <- telco_cleaned %>%
  count(Churn_Label) %>%
  mutate(percentage = round(n / sum(n) * 100, 1))
# Plot churn distribution
ggplot(churn_rate, aes(x = factor(Churn_Label), y = n, 
                       fill = as.factor(Churn_Label))) +
    geom_col() +
    labs(title = "Churn Distribution for all Customers", y = "Count") +
    scale_x_discrete(labels = c("0" = "Did Not Churn", "1" = "Churned")) + 
    geom_text(aes(label = paste0(percentage, "%")),
              position = position_stack(vjust = 0.5)) +
    theme(legend.position = "none", axis.title.x = element_blank())
# How many are high value customers
high_value <- telco_cleaned %>%
  filter(CLTV > median(CLTV)) %>% # filter based on the median value of CLTV
  group_by(Churn_Label) %>%
  summarize(Count = n()) %>%
  mutate(percentage = round(Count / sum(Count) * 100, 1))
# Plot high value customer distribution
ggplot(high_value, aes(x = factor(Churn_Label), y = Count, 
                       fill = as.factor(Churn_Label))) +
    geom_col() +
    labs(title = "Churn Distribution for High value Customers", y = "Count") +
    geom_text(aes(label = paste0(percentage, "%")),
              position = position_stack(vjust = 0.5)) +
    scale_x_discrete(labels = c("0" = "Did Not Churn", "1" = "Churned")) +
    theme(legend.position = "none", axis.title.x = element_blank())
```

`r churn_rate[2, 2]` customers out of which `r high_value[2, 2]` are high value customers have churned. 
A churn rate of `r high_value[2, 3]`% for high value customers and overall churn rate of `r churn_rate[2, 3]`% for all customers, both are considerably high.

Next, it is important to find which variables have high correlation with churn.

```{r churn_correlation, fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
# Correlation matrix
cor_matrix <- cor(telco_cleaned %>%
                    select_if(is.numeric))
corrplot(cor_matrix, method = "color",
         addCoef.col = "black", tl.col = "red", 
         number.font = 0.3, number.digits = 1)
```

Correlation plot showing the relationship between the variables.Tenure Months
and Total Charges have the strongest relationship, but other relationships do
exists, which should help in the analysis.

```{r feat_importance, warning=FALSE, message=FALSE}
telco_split <- initial_split(telco_cleaned, prop = 0.7, strata = Churn_Label)
train_data <- training(telco_split)
test_data <- testing(telco_split)
# Random Forest model for feature importance
rf_model <- randomForest(Churn_Label ~ ., data = train_data)
varImpPlot(rf_model)
```

The Random Forest model provides insight to those variables that contribute
to customer churn. The model allows us to focus on and analyze some of these key
predictors that strongly influence customer churn.

## 2. Can we identify patterns or behaviors that precede customer churn?

To do that, the data needs to only include churned customers, then further analysis could identify which factors are unique to them. Specifically, we look at those variable that are high on the random forest model.

```{r churn_distribution, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}
# Create a subset of data for customers who churned
churned_customers <- telco_cleaned %>%
    filter(Churn_Label == 1)
# Analyze Contract Type by churn
d1 <- ggplot(churned_customers, aes(x = Contract, fill = Contract)) +
        geom_bar() +
        labs(title = "Contract Type for Churned Customers", 
             x = "Contract", y = "Count") +
        theme(legend.position = "none")
# Analyze Tenure Months by churn
d2 <- ggplot(churned_customers, aes(x = Tenure_Months)) +
        geom_histogram(binwidth = 1,
                       fill = "blue",
                       color = "black") +
        labs(title = "Tenure Distribution for Churned Customers",
             x = "Tenure in Months", y = "Count")
# Analyze Internet Service Type by churn
d3 <- ggplot(churned_customers,
             aes(x = Internet_Service, fill = Internet_Service)) +
        geom_bar() +
        labs(title = "Internet Service Type for Churned Customers", 
             x = "Internet Service", y = "Count") +
        theme(legend.position = "none")
# Analyze Payment Method Type by churn
d4 <- ggplot(churned_customers,
       aes(x = Payment_Method, fill = Payment_Method)) +
    geom_bar() +
    labs(title = "Payment Method Type for Churned Customers", 
         x = "Payment Method", y = "Count") +
    scale_x_discrete(labels=c("Bank Transfer\n(automatic)", 
                                "Credit Card\n(automatic)",
                                "Electronic Check",
                               "Mailed Check")) +
    theme(legend.position = "none") 
# Print charts
grid.arrange(d1, d2, d3, d4, ncol = 2)
```

Looking at the charts above, we can deduce that most customers who churn have a month-to-month contract and tend to stay for short periods because the tenure distribution is right-skewed, they also subscribed to the fibre-optic internet service and chose to pay by electronic check. 
These are certainly behaviors that could be used to track churning.

## 3. Which customer segments have the highest churn rates?

Analyse the demography of churned customers by segmenting to look at the gender,
those classified as senior citizens and those living with partners and
dependents to find any patterns.

```{r, churn_by_segment, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}
# Calculate Churn Rate by Gender
churn_rate_by_Gender <- telco_cleaned %>%
    group_by(Gender) %>%
    summarize(Churn_Rate = mean(Churn_Label))
# Plot Churn Rate by Gender
s1 <- ggplot(churn_rate_by_Gender,
             aes(x = Gender, y = Churn_Rate, fill = Gender)) +
          geom_col() +
          geom_text(aes(label = paste0(round(Churn_Rate, 2))),
              position = position_stack(vjust = 0.5)) +
          labs(title = "Churn Rate by Gender", 
               x = "Gender", y = "Churn Rate") +
          theme(legend.position = "top")
# Calculate Churn Rate by Senior Citizen
churn_rate_by_Age <- telco_cleaned %>%
    group_by(Senior_Citizen) %>%
    summarize(Churn_Rate = mean(Churn_Label))
# Plot Churn Rate by Senior Citizen
s2 <- ggplot(churn_rate_by_Age,
             aes(x = as.factor(Senior_Citizen), y = Churn_Rate, 
                 fill = as.factor(Senior_Citizen))) +
        geom_col() +
          geom_text(aes(label = paste0(round(Churn_Rate, 2))),
              position = position_stack(vjust = 0.5)) +
        labs(title = "Churn Rate by Senior Citizen",
             x = "Senior Citizen", y = "Churn Rate") +
        scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
        scale_fill_manual(values = c("0" = "blue", "1" = "red"),
                          labels = c("0" = "No", "1" = "Yes"),
                          name = "Senior Citizen") +
        theme(legend.position = "top")
# Calculate Churn Rate by those living with Dependents
churn_rate_by_Dependents <- telco_cleaned %>%
    group_by(Dependents) %>%
    summarize(Churn_Rate = mean(Churn_Label))
# Plot Churn Rate by those living with Dependents
s3 <- ggplot(churn_rate_by_Dependents,
             aes(x = as.factor(Dependents), y = Churn_Rate, 
                 fill = as.factor(Dependents))) +
          geom_col() +
          geom_text(aes(label = paste0(round(Churn_Rate, 2))),
              position = position_stack(vjust = 0.5)) +
          labs(title = "Churn Rate by Dependents",
               x = "Living with Dependents", y = "Churn Rate") +
          scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
          scale_fill_manual(values = c("0" = "blue", "1" = "red"),
                            labels = c("0" = "No", "1" = "Yes"),
                            name = "Living with Dependents") +
          theme(legend.position = "top")
# Calculate Churn Rate by Gender
churn_rate_by_Partner <- telco_cleaned %>%
    group_by(Partner) %>%
    summarize(Churn_Rate = mean(Churn_Label))
# Plot Churn Rate by those with a Partner
s4 <- ggplot(churn_rate_by_Partner,
             aes(x = as.factor(Partner), y = Churn_Rate, 
                 fill = as.factor(Partner))) +
          geom_col() +
          geom_text(aes(label = paste0(round(Churn_Rate, 2))),
              position = position_stack(vjust = 0.5)) +
          labs(title = "Churn Rate by Partner",
               x = "Living with Partner", y = "Churn Rate") +
          scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
          scale_fill_manual(values = c("0" = "blue", "1" = "red"),
                            labels = c("0" = "No", "1" = "Yes"),
                            name = "Living with a Partner") +
          theme(legend.position = "top")
# Print charts
grid.arrange(s1, s2, s3, s4, ncol = 2)
```

It looks like gender has no impact on churn rate, while Senior Citizens and
those living without dependents or partners are more likely to churn. Efforts to
retain more senior citizens and those living without any dependents or partners
should be the focus.

## 4. Are there specific products or services associated with higher churn rates?

It is time to look at the churn rates for specific services, to find any
patterns.

```{r, churn_by_service, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}
churn_rate_by_Phone_Service <- telco_cleaned %>%
    group_by(Phone_Service) %>%
    summarize(Churn_Rate = mean(Churn_Label))
churn_rate_by_Multiple_Lines <- telco_cleaned %>%
    group_by(Multiple_Lines) %>%
    summarize(Churn_Rate = mean(Churn_Label))
churn_rate_by_Online_Security <- telco_cleaned %>%
    group_by(Online_Security) %>%
    summarize(Churn_Rate = mean(Churn_Label))
churn_rate_by_Online_Backup <- telco_cleaned %>%
    group_by(Online_Backup) %>%
    summarize(Churn_Rate = mean(Churn_Label))
churn_rate_by_Device_Protection <- telco_cleaned %>%
    group_by(Device_Protection) %>%
    summarize(Churn_Rate = mean(Churn_Label))
churn_rate_by_Tech_Support <- telco_cleaned %>%
    group_by(Tech_Support) %>%
    summarize(Churn_Rate = mean(Churn_Label))
churn_rate_by_Streaming_TV <- telco_cleaned %>%
    group_by(Streaming_TV) %>%
    summarize(Churn_Rate = mean(Churn_Label))
churn_rate_by_Streaming_Movies <- telco_cleaned %>%
    group_by(Streaming_Movies) %>%
    summarize(Churn_Rate = mean(Churn_Label))
churn_rate_by_Paperless_Billing <- telco_cleaned %>%
    group_by(Paperless_Billing) %>%
    summarize(Churn_Rate = mean(Churn_Label))
# Create a bar plot to visualize churn rates by those with phone service
p1 <- ggplot(churn_rate_by_Phone_Service, 
             aes(x = as.factor(Phone_Service), y = Churn_Rate, 
                 fill = as.factor(Phone_Service))) +
    geom_col() +
  geom_text(aes(label = paste0(round(Churn_Rate, 2))),
              position = position_stack(vjust = 0.5)) +
    labs(title = "Churn Rates by Phone Service", 
         x = "Have phone service", 
         y = "Churn Rate") +    
  scale_fill_manual(values = c("0" = "blue", "1" = "red"),
                      labels = c("0" = "No", "1" = "Yes")) +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    theme(legend.position = "none")
# Create a bar plot to visualize churn rates by those with multiple lines
p2 <- ggplot(churn_rate_by_Multiple_Lines, 
             aes(x = as.factor(Multiple_Lines), y = Churn_Rate, 
                 fill = as.factor(Multiple_Lines))) +
    geom_col() +
  geom_text(aes(label = paste0(round(Churn_Rate, 2))),
              position = position_stack(vjust = 0.5)) +
    labs(title = "Churn Rates by Multiple Lines", 
         x = "Have multiple lines", 
         y = "Churn Rate") +    
  scale_fill_manual(values = c("0" = "blue", "1" = "red"),
                      labels = c("0" = "No", "1" = "Yes")) +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    theme(legend.position = "none")
# Create a bar plot to visualize churn rates by those with online security
p3 <- ggplot(churn_rate_by_Online_Security, 
             aes(x = as.factor(Online_Security), y = Churn_Rate, 
                 fill = as.factor(Online_Security))) +
    geom_col() +
  geom_text(aes(label = paste0(round(Churn_Rate, 2))),
              position = position_stack(vjust = 0.5)) +
    labs(title = "Churn Rates by Online Security", 
         x = "Have online security", 
         y = "Churn Rate") +    
  scale_fill_manual(values = c("0" = "blue", "1" = "red"),
                      labels = c("0" = "No", "1" = "Yes")) +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    theme(legend.position = "none")
# Create a bar plot to visualize churn rates by those with online backup
p4 <- ggplot(churn_rate_by_Online_Backup, 
             aes(x = as.factor(Online_Backup), y = Churn_Rate, 
                 fill = as.factor(Online_Backup))) +
    geom_col() +
  geom_text(aes(label = paste0(round(Churn_Rate, 2))),
              position = position_stack(vjust = 0.5)) +
    labs(title = "Churn Rates by Online Backup", 
         x = "Have online backup", 
         y = "Churn Rate") +    
  scale_fill_manual(values = c("0" = "blue", "1" = "red"),
                      labels = c("0" = "No", "1" = "Yes")) +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    theme(legend.position = "none")
# Create a bar plot to visualize churn rates by those with device protection
p5 <- ggplot(churn_rate_by_Device_Protection, 
             aes(x = as.factor(Device_Protection), y = Churn_Rate, 
                 fill = as.factor(Device_Protection))) +
    geom_col() +
  geom_text(aes(label = paste0(round(Churn_Rate, 2))),
              position = position_stack(vjust = 0.5)) +
    labs(title = "Churn Rates by Device Protection", 
         x = "Have device protection", 
         y = "Churn Rate") +    
  scale_fill_manual(values = c("0" = "blue", "1" = "red"),
                      labels = c("0" = "No", "1" = "Yes")) +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    theme(legend.position = "none")
# Create a bar plot to visualize churn rates by those with tech support
p6 <- ggplot(churn_rate_by_Tech_Support, 
             aes(x = as.factor(Tech_Support), y = Churn_Rate, 
                 fill = as.factor(Tech_Support))) +
    geom_col() +
  geom_text(aes(label = paste0(round(Churn_Rate, 2))),
              position = position_stack(vjust = 0.5)) +
    labs(title = "Churn Rates by Tech Support", 
         x = "Have tech support", 
         y = "Churn Rate") +    
  scale_fill_manual(values = c("0" = "blue", "1" = "red"),
                      labels = c("0" = "No", "1" = "Yes")) +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    theme(legend.position = "none")
# Create a bar plot to visualize churn rates by that stream TV
p7 <- ggplot(churn_rate_by_Streaming_TV, 
             aes(x = as.factor(Streaming_TV), y = Churn_Rate, 
                 fill = as.factor(Streaming_TV))) +
    geom_col() +
  geom_text(aes(label = paste0(round(Churn_Rate, 2))),
              position = position_stack(vjust = 0.5)) +
    labs(title = "Churn Rates by Streaming TV", 
         x = "Stream TV", 
         y = "Churn Rate") +    
  scale_fill_manual(values = c("0" = "blue", "1" = "red"),
                      labels = c("0" = "No", "1" = "Yes")) +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    theme(legend.position = "none")
# Create a bar plot to visualize churn rates by that stream Movies
p8 <- ggplot(churn_rate_by_Streaming_Movies, 
             aes(x = as.factor(Streaming_Movies), y = Churn_Rate, 
                 fill = as.factor(Streaming_Movies))) +
    geom_col() +
  geom_text(aes(label = paste0(round(Churn_Rate, 2))),
              position = position_stack(vjust = 0.5)) +
    labs(title = "Churn Rates by Streaming Movies", 
         x = "Stream movies", 
         y = "Churn Rate") +    
  scale_fill_manual(values = c("0" = "blue", "1" = "red"),
                      labels = c("0" = "No", "1" = "Yes")) +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    theme(legend.position = "none")
# Create a bar plot to visualize churn rates by that subscribed to Paperless billing
p9 <- ggplot(churn_rate_by_Paperless_Billing, 
             aes(x = as.factor(Paperless_Billing), y = Churn_Rate, 
                 fill = as.factor(Paperless_Billing))) +
    geom_col() +
  geom_text(aes(label = paste0(round(Churn_Rate, 2))),
              position = position_stack(vjust = 0.5)) +
    labs(title = "Churn Rates by Paperless Billing", 
         x = "Subscribed to paperless billing", 
         y = "Churn Rate") +    
  scale_fill_manual(values = c("0" = "blue", "1" = "red"),
                      labels = c("0" = "No", "1" = "Yes")) +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    theme(legend.position = "none")
# Print charts
grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, ncol = 3)
```

The charts above show the services that have high churn rates. Customers that
don't subscribe to have Online Security, tech support and those that subscribe
to have paperless billing are more likely to churn.
By addressing customer pain points associated with these services, the rate of churn could be reduced overall, as well as increasing customer satisfaction.

## 5. Can we predict which customers are likely to churn in the near future?

This would involve building machine learning algorithms that could help predict customers that would churn. Due to nature of the data, a logistic regression model and a classification decision tree model will be built and evaluated to select the best among the two.

```{r, churn_regression, fig.height=6, fig.width=6, warning=FALSE, message=FALSE}
# Set seed for reproducibility
set.seed(123)
# Split data using initial_split() 70% for training, 30% for testing
data_split <- initial_split(telco_cleaned, prop = 0.7, strata = Churn_Label)
train_data <- training(data_split)
test_data <- testing(data_split)
# Build a logistic regression model for probabilities
churn_model <- glm(Churn_Label ~ .,
                   data = train_data,
                   family = "binomial")
# Predict churn probabilities on the test set
churn_probabilities <- predict(churn_model, 
                               newdata = test_data, 
                               type = "response")
# Evaluate model performance using ROC curve
roc_log <- roc(test_data$Churn_Label, churn_probabilities)
plot(roc_log, main = "ROC Curve for Logistic Regression Model")
auc_log <- auc(roc_log)
# Set threshold value
threshold <- 0.5
# Convert churn probabilities to binary predictions (0 or 1) based on threshold value set above
predicted_churn <- ifelse(churn_probabilities >= threshold, 1, 0)
# Compare predicted churn with actual churn in the test set
confusion_matrix <- table(predicted_churn, test_data$Churn_Label)
# Calculate accuracy, precision, and recall using confusion matrix
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
# Precision = True Positives / (True Positives + False Positives)
precision <- confusion_matrix[2, 2] / sum(confusion_matrix[, 2])
# Recall = True Positives / (True Positives + False Negatives)
recall <- confusion_matrix[2, 2] / sum(confusion_matrix[2, ])      
# Print the model performance measurement metrics
cat("For the Logistic Regression Model, the performance metrics are: ", "\n",
"Accuracy: ", round(accuracy, 3) * 100, "%", "\n",
"Precision: ", round(precision, 3) * 100, "%", "\n",
"Recall: ", round(recall, 3)* 100, "%", "\n",
"Area Under Curve: ", round(auc_log, 3))
```

AUC (Area Under Curve) for the Logistic Regression Model is `r round(auc_log, 3)`, which indicates the model is a good fit.
Other performance metrics for the Logistic Regression Model are:
Accuracy of `r paste0(round(accuracy, 3) * 100)`%, 
Precision of `r paste0(round(precision, 3) * 100)`%, and
Recall of `r paste0(round(recall, 3) * 100)`%, which all indicate the model performance is quite good and acceptable.

```{r, churn_classification, fig.height=6, fig.width=6, warning=FALSE, message=FALSE}
# Set seed for reproducibility
set.seed(123)
# Build a Decision Tree model for classification
churn_tree <- rpart(Churn_Label ~ ., 
                    data = train_data, 
                    method = "class")
# Predict probabilities on the test set
churn_probabilities <- predict(churn_tree, 
                               newdata = test_data, 
                               type = "prob")[, 2]
# Evaluate model performance using ROC curve
roc_DT <- roc(test_data$Churn_Label, churn_probabilities)
plot(roc_DT, main = "ROC Curve for Decision Tree Model")
auc_DT <- auc(roc_DT)
# Evaluate model performance using a confusion matrix
tree_confusion_matrix <- table(churn_probabilities, test_data$Churn_Label)
# Calculate accuracy = True Positives / (All Observations)
accuracy2 <- sum(diag(tree_confusion_matrix)) / sum(tree_confusion_matrix)
# Calculate Precision = True Positives / (True Positives + False Positives)
precision2 <- tree_confusion_matrix[2, 2] / sum(tree_confusion_matrix[, 2])
# Calculate Recall = True Positives / (True Positives + False Negatives)
recall2 <- tree_confusion_matrix[2, 2] / sum(tree_confusion_matrix[2, ])      
# Print the model performance measurement metrics
cat("For the Decision Tree Model, the performance metrics are: ", "\n",
"Accuracy: ", round(accuracy2, 3) * 100, "%", "\n",
"Precision: ", round(precision2, 3) * 100, "%", "\n",
"Recall: ", round(recall2, 3)* 100, "%", "\n",
"Area Under Curve: ", round(auc_DT, 3))
```

AUC (Area Under Curve) for the Decision Tree Model is `r round(auc_DT, 3)`, which indicates the model is also a good fit.
Other performance metrics for the Decision Tree Model are:
Accuracy of `r paste0(round(accuracy2, 3) * 100)`%, 
Precision of `r paste0(round(precision2, 3) * 100)`%, and
Recall of `r paste0(round(recall2, 3) * 100)`%, which all indicate a poor and unacceptable performance, even though the fit is good.

#### Therefore, in order to predict which customers will churn in the future, the logistic regression model is the model of choice.

# Limitations

1. The analysis couldn't consider the customer reasons for churning due to the high proportion of missing values for that variable, and as such, the insights into the specific motivations or pain points driving churn remain limited. This missing information could potentially lead to an incomplete understanding of the underlying factors contributing to customer attrition and might hinder the development of targeted strategies to address these issues effectively.

2. While the analysis identifies correlations between the variables and churn,
it is important to remember correlation isn't causation, and therefore a lot of
other factors could be responsible for churn. For example, the data does not consider temporal trends that could affect churn rates over time. 

3. Logistic regression and decision tree models make certain assumptions about
data distribution and relationships, and any violation of these assumptions
could affect the validity of the analysis.

# Conclusion

1. It is important to improve customer retention strategies that target the
specific customers with high churn rate through personalized communication,
improved customer service and ensuring the quality and reliability of services,
leading to an improved customer satisfaction rate which could result in a low
churn rate.

2. The Offering of loyalty programs such as rewards, discounts, and exclusive benefits to long-term customers can incentivize them to stay and continue using your products or service, while also setting competitive prices that are affordable can influence customers' to stay.

3. Implementing an early warning systems such as a predictive model that can
identify customers at risk of churning, to allow proactive interventions to
prevent churn. This could also be achieve with a strong customer feedback system
that offer suggestions on improvements and pain points.

4. It is important to implement an exit survey conducted for churned customers
to include not only reasons for leaving, but also customer satisfaction ratings
on services can provide insights into why they left, to guide strategies to
address common pain points.