---
title: "Airline Delay Prediction"
author: "Aliyu Atiku Mustapha"
date: "2023-09-11"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 2
    code_folding: hide
---

# Introduction

### The task is to build a regression model that can predict whether a given flight will be delayed, given the information of the scheduled departure.

# Data

## Airline Delay Data

The data contains 539383 measurements of 9 different attributes.

The dataset is available in the [UCI Machine Learning
Repository:](https://archive.ics.uci.edu/dataset/477/real+estate+valuation+data+set).

| Column Name | Column Description 
|--------- |--------- 
| id    |  ID column
| Airline    | Airline Name
| Flight    | Flight Number
| AirportFrom    | Departure Airport
| AirportTo    | Arrival Airport
| DayofWeek    | Day of week of flight
| Time | Time of flight
| Length | Length of delay
| Delay | Delayed or not

```{r setup, warning=FALSE, message=FALSE}
# Load tidyverse package
library(tidyverse)
# Load rsample package for correlation plot
library(rsample)
# Load corrplot package for correlation plot
library(corrplot)
# Load gbm package for Logistic regression modeling
library(gbm)
# Load randomForest package for Random Forest model
library(randomForest)
# Load pROC package for ROC curve
library(pROC)
# Load DT package
library(DT)
# Load knitr package
library(knitr)
# Set constant size for plots
knitr::opts_chunk$set(fig.width = 8, fig.height = 5)
```


## Data Loading
```{r read_data_file, message=FALSE, warning=FALSE}
airline_data <- read.csv("E:/My Data Science Portfolio/Datasets/Airlines.csv")
# Calculate total number of rows in data
cat("The Telco dataset has",nrow(airline_data), "rows.")
# Display the first 500 rows of imported data
DT::datatable(
  airline_data[1:500,],
  options = list(pageLength = 20),
  fillContainer = T,
  caption = "Table displaying first 500 rows of loaded data.")
```

## Data Properties

To understand the data composition, the class of each variable together with
the number and proportion of missing values for each variable will provide a 
deeper insight to the data structure and how it could be useful for analysis.

```{r data_str, warning=FALSE, message=FALSE}
# Check for class of each variable
class_table <- sapply(airline_data, class)
class_table <- data.frame(
  Variable = names(class_table),
  Class = as.character(class_table),
  stringsAsFactors = FALSE
)
# Check for the proportion of missing values in full data
x <- data.frame(miss = colSums(is.na(airline_data)),
                prop = round((colSums(is.na(airline_data)) 
                              / nrow(airline_data)) * 100,
                          2))
properties <- cbind(class_table,
                    Values_missing = x[, 1],
                    Proportion_missing = x[, 2])
# Display the properties of the data
kable(head(properties, 10),
      caption = "Table displaying the properties of data.")
```

No missing data, no need for cleaning or transformation, so dive straight to understand the data.

# Analysis

The objective of this analysis is to understand the flight data.

## Number of flights delayed vs not delayed

```{r delayed_vs_undelayed, message=FALSE, warning=FALSE}
# Visualize number of flights delayed and not delayed
ggplot(airline_data, aes(x = as.factor(Delay), 
                               fill = as.factor(Delay))) +
    geom_bar() +
    labs(title = "Distribution of flight delay", 
         x = "Flight Delayed", y = "Count") +
    theme(legend.position = "top") + 
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
    scale_fill_manual(values = c("0" = "blue", "1" = "red"),
                      labels = c("0" = "No", "1" = "Yes"),
                      name = "Flight Delayed")
```

There is a small difference between the number of total flights delayed and those not delayed.

## Number of delays on each Day

```{r daily_analysis, message=FALSE, warning=FALSE}
# Visualize number of flights delayed on each day
airline_data %>%
    filter(Delay == "1") %>%
    group_by(DayOfWeek) %>%
    count(Delay) %>%
    ggplot(aes(x = DayOfWeek, y = n, fill = as.factor(DayOfWeek))) +
    geom_col() +
    geom_text(
    aes(label = paste0(n)),
    position = position_dodge(width = 1), vjust = -0.3) +
        labs(title = "Distribution delays by day",
             x = "Day of Week", y = "Count") +
    theme(legend.position = "none")
```

The distribution of delays by day indicates mid-week days have a higher number of flight delays, while Saturdays have the least.

## Number of delays per Airline

```{r airline_analysis, message=FALSE, warning=FALSE}
# Airlines with highest number of flight delays
airline_data %>%
  group_by(Airline) %>%
  summarize(
    Total_flights = n(),
    delayed_flights = sum(Delay),
    prop_delayed = round((delayed_flights / Total_flights) * 100, 0)) %>%
  ggplot(aes(x = reorder(Airline, -prop_delayed))) +
  geom_bar(aes(y = Total_flights, fill = "Total_flights"),
           stat = "identity", position = "dodge") +
  geom_bar(
    aes(y = delayed_flights, fill = "delayed_flights"),
    stat = "identity", position = "dodge") +
  geom_text(
    aes(y = Total_flights, label = paste0(prop_delayed, "%")),
    position = position_dodge(width = 1), vjust = -0.3) +
  labs(x = "Airline", y = "Count", 
       title = "Total Flights vs. Delayed Flights by Airline",
       subtitle = "Prepare for delays!") +
  scale_fill_manual(
    values = c("Total_flights" = "blue", "delayed_flights" = "red"),
    labels = c("Total_flights" = "Total Flights", 
               "delayed_flights" = "Delayed Flights"))
```

Customers flying on WN Airline and CO Airline are more likely to be delayed while other airlines such as B6, DL, F9 and OO also tend to experience delays frequently.

## Departure Airports that are more likely to delay

```{r departure_analysis, message=FALSE, warning=FALSE}
# Visualize 30 Departure Airports with highest number of Delays
airline_data %>%
  group_by(AirportFrom) %>%
  summarize(
    Total_flights = n(),
    delayed_flights = sum(Delay),
    prop_delayed = round((delayed_flights / Total_flights) * 100, 0)
  ) %>%
  filter(prop_delayed > 50) %>%
  slice_max(order_by = prop_delayed, n = 30)  %>%
  arrange(desc(prop_delayed)) %>%
  ggplot(aes(x = reorder(AirportFrom,-prop_delayed))) +
  geom_bar(aes(y = Total_flights, fill = "Total_flights"),
           stat = "identity", position = "dodge") +
  geom_bar(
    aes(y = delayed_flights, fill = "delayed_flights"),
    stat = "identity", position = "dodge") +
  geom_text(
    aes(y = Total_flights, label = paste0(prop_delayed, "%")),
    position = position_dodge(width = 1),
    vjust = -0.3,
    angle = 45) +
  labs(x = "Departure Airport", y = "Count", 
       title = "Total Flights vs. Delayed Flights by Departure Airport",
       subtitle = "Will my flight depart on time?") +
  scale_fill_manual(
    values = c("Total_flights" = "blue",
               "delayed_flights" = "red"),
    labels = c("Total_flights" = "Total Flights", 
               "delayed_flights" = "Delayed Flights")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Airports such as MDW, DAL, OAK, HOU and OTH are top 5 airports with high proportion of flight delays.

## Arrival Airports that are more likely to delay

```{r arrival_analysis, message=FALSE, warning=FALSE}
# Visualize 30 Arrival Airports with highest number of Delays
airline_data %>%
    group_by(AirportTo) %>%
    summarize(
        Total_flights = n(),
        delayed_flights = sum(Delay),
        prop_delayed = round((delayed_flights / Total_flights) * 100, 0)) %>%
  filter(prop_delayed > 50) %>%
  slice_max(order_by = prop_delayed, n = 30)  %>%
  arrange(desc(prop_delayed)) %>%
  ggplot(aes(x = reorder(AirportTo,-prop_delayed))) +
  geom_bar(aes(y = Total_flights, fill = "Total_flights"),
           stat = "identity", position = "dodge") +
  geom_bar(
    aes(y = delayed_flights, fill = "delayed_flights"),
    stat = "identity", position = "dodge") +
  geom_text(
    aes(y = Total_flights, label = paste0(prop_delayed, "%")),
    position = position_dodge(width = 1), 
    vjust = -0.3,
    angle = 45) +
  labs(x = "Arrival Airport", y = "Count", 
       title = "Total Flights vs. Delayed Flights by Arrival Airport",
       subtitle = "Will my flight arrive on time?") +
  scale_fill_manual(
    values = c("Total_flights" = "blue", 
               "delayed_flights" = "red"),
    labels = c("Total_flights" = "Total Flights", 
               "delayed_flights" = "Delayed Flights")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Airports such as OAK, ISP, MDW, ONT and RDD are the top 5 airports with high proportion of delays.

Customers using airports such as MDW, DAL, OAK, HOU and SMF are bound to face delays departing or arriving.

# Model

First, it is important to find which variables have high linear correlation with delay, if any.

## Correlation Plot

```{r churn_correlation, warning=FALSE, message=FALSE}
# Correlation matrix
cor_matrix <- cor(airline_data %>%
                    select_if(is.numeric))
corrplot(cor_matrix, method = "color",
         addCoef.col = "black", tl.col = "red")
```

From the correlation result, Delay lacks strong linear relationships with any of the variables in the data, which makes choosing Logistic Regression easy for the model type because 'Logistic regression' predicts on binary outcomes such as "YES" or "NO" and assumes nothing about the relationship type.

## Split data into training and testing sets

```{r split_data, message=FALSE, warning=FALSE}
# Set seed for reproducibility
set.seed(123)
# Split data, 70% for training, 30% for testing
airline_split <-
  initial_split(airline_data, prop = 0.7, strata = Delay)
train_data <- training(airline_split)
test_data <- testing(airline_split)
```

The data has been split into a training data containing approximately 70% of the data and test data containing approximately 30% of the data, next step involves building the logistic model using the training data.

## Logistic Regression Model

```{r logreg_model, message=FALSE, warning=FALSE}
# Build and fit a logistic regression model
delay_model <- glm(Delay ~ ., data = train_data, family = "binomial")

# Predict target_variable probabilities on the test set
delay_probabilities <- predict(delay_model, 
                                         newdata = test_data, 
                                         type = "response")

# Evaluate model performance using ROC curve
roc <- roc(test_data$Delay, delay_probabilities)
plot(roc, main = "ROC Curve")
auc_roc <- auc(roc)
# Calculate Area Under the Curve for the model
cat("The Area Under the Curve (AUC) for the Logistic Regression model is",
    auc_roc, "\n")
```

The AUC indicates this is a good model but the fit is not excellent.

Let see how the model predicts and measure its performance using other metrics such as 'Accuracy', 'Precision' and 'Recall'.

```{r logreg_metrics, message=FALSE, warning=FALSE}
# Set threshold value
threshold <- 0.5
# Convert target_variable probabilities to binary predictions (0 or 1) based on threshold
predicted_delay <- ifelse(delay_probabilities >= threshold, 1, 0)

# Compare predicted variable with actual variable in the test set
logreg_confusion_matrix <- table(predicted_delay, test_data$Delay)
# Calculate accuracy, precision, and recall using confusion matrix
# Accuracy = True Positives / (Observations Predicted Correctly)
logreg_accuracy <-
  sum(diag(logreg_confusion_matrix)) / sum(logreg_confusion_matrix)
# Precision = True Positives / (All Observations)
logreg_precision <-
  logreg_confusion_matrix[2, 2] / sum(logreg_confusion_matrix[, 2])
# Recall = True Positives / (True Positives + False Negatives)
logreg_recall <-
  logreg_confusion_matrix[2, 2] / sum(logreg_confusion_matrix[2,])

# Print the metrics
cat("The Logistic Regression model has the following metrics: ", "\n",
    "Accuracy: ", round(logreg_accuracy, 2) * 100, "%", "\n",
    "Precision:", round(logreg_precision, 2) * 100, "%","\n",
    "Recall:", round(logreg_recall, 2) * 100, "%")
```

The performance of the model can best be described as average.
Since the model is a logic regression model that depends on a threshold value for the conversion of probabilities to a binary outcome, the threshold value can be varied to see how that affects the performance of the model.

## Improve Model

The threshold will vary from 0.1 to 0.9 in steps of 0.1, to see if a best value can be found.

```{r tune_logreg_model, message=FALSE, warning=FALSE}
# Define a sequence of threshold values
thresholds <- seq(0.1, 0.9, by = 0.1)
results <- list()
for (threshold in thresholds) {
  # Predict target_variable probabilities on the test set
  delay_probabilities <- predict(delay_model,
                                 newdata = test_data,
                                 type = "response")
  # Convert target_variable probabilities to binary predictions (0 or 1) based on threshold
  predicted_delay <- ifelse(delay_probabilities >= threshold, 1, 0)
  # Calculate confusion matrix and metrics
  tuned_logreg_confusion_matrix <-
    table(predicted_delay, test_data$Delay)
  logreg_accuracy <-
    sum(diag(tuned_logreg_confusion_matrix)) / sum(tuned_logreg_confusion_matrix)
  logreg_precision <-
    tuned_logreg_confusion_matrix[2, 2] / sum(tuned_logreg_confusion_matrix[, 2])
  logreg_recall <-
    tuned_logreg_confusion_matrix[2, 2] / sum(tuned_logreg_confusion_matrix[2, ])
  
  results[[as.character(threshold)]] <- list(
    tuned_accuracy = round(logreg_accuracy, 2) * 100,
    tuned_precision = round(logreg_precision, 2) * 100,
    tuned_recall = round(logreg_recall, 2) * 100
  )
  # Print the metrics for the threshold value
  cat(
    "For threshold set to", threshold,"-\n",
    "Accuracy:", results[[as.character(threshold)]]$tuned_accuracy, "\n",
    "Precision:", results[[as.character(threshold)]]$tuned_precision, "\n",
    "Recall:", results[[as.character(threshold)]]$tuned_recall, "\n")
}
```

# Conclusion

A Logistic Regression model is not a good fit for predicting delays based on the
data provided, the Accuracy, Precision and Recall of the model are at best considered average,
despite trying different threshold values for converting the probabilities of delay to a binary
outcome.
This suggests that a logistic Regression model will not be the best model of
choice for this task, and another model will have to be tried, (probably a
decision tree or ensemble model).